---
title: Simulated Pre- and Post-Test Data
---

## Dataset Description 

A simulated dataset used for various data analysis examples in teaching. The dataset contains three variables:

- `Y1` (pre-test scores),
- `X` (treatment assignment), and
- `Y2` (post-test scores). 


You can download the data as a `.rds` file here: [sim_pre_post_data.rds](https://riegersven.github.io/library/example-data/sim_pre_post_data.rds), or generate it yourself using the function in @lst-sim-pre-post-data.


## Data Generating Process

The data generating process is based on a simple pre-post study design with a binary treatment variable, as illustrated in [@fig-pre-post]. The pre-test scores (`Y1`) are generated from a standard normal distribution. The treatment assignment (`X`) is generated from a Bernoulli distribution with a specified probability of treatment. The post-test scores (`Y2`) are generated based on a linear model that includes the treatment effect (`tau`), the effect of the pre-test scores (`b1`), and an error term with a specified residual variance. 

::: {.column-margin}

```{mermaid}
%%| label: fig-pre-post
%%| echo: false
%%| fig-cap: "An example pre-post study design"
%%| fig-align: left

flowchart LR
  Y1 --> R
  R --.5--> T
  R --.5--> C
  T --> Y2
  C --> Y2
```

:::

::: {#lst-sim-pre-post-data}

```{r}
#| label: sim-pre-post-data
#| code-fold: show
sim_pre_post_data <- function ( n = 500,
                                treat_prob = 0.50,
                                b0 = 0,
                                tau = 0.25,
                                b1 = 0.50,
                                b2 = 0,
                                seed = 42 ) {
  set.seed(seed)
  
  Y1 <- rnorm(n = n,
              mean = 0,
              sd = 1)

  X <- rbinom(n = n,
              size = 1,
              prob = treat_prob)   

  XY1 <- X*Y1
              
  resid_var <- 1 - (tau^2 * var(X) + 
                    b1^2 * var(Y1) +
                    b2^2 * var(XY1) +
                    2 * tau * b1 * cov(X, Y1) + 
                    2 * tau * b2 * cov(X, XY1) +
                    2 * b1 * b2 * cov(Y1, XY1)
                )

  if (resid_var < 0) {
    stop("The specified parameters lead to a negative residual variance. Please adjust the parameters.")
  }

  Y2 <- b0 + tau*X + b1*Y1 + 
        rnorm(n = n,
              mean = 0,
              sd = sqrt(resid_var))

  ret <- data.frame(Y1 = Y1,
                    X = X,
                    Y2 = Y2)
  ret
}

```

:::


The function `sim_pre_post_data` allows you to specify the number of observations (`n`), the probability of treatment assignment (`treat_prob`), the intercept (`b0`), the treatment effect (`tau`), the effect of the pre-test scores (`b1`), the effect of the interaction between treatment and pre-test scores (`b2`), and a random seed for reproducibility. Without any input arguments, the function generates a dataset with the default parameters specified above. 

```{r}
#| label: exe-sim_pre_post_data
#| output-location: fragment

ex_dat <- sim_pre_post_data()
head(ex_dat)
```



```{r}
#| label: sim-pre-post-data-export
#| echo: false
#| eval: false
saveRDS(ex_dat, file = "example-data/sim_pre_post_data.rds")
```