---
title: Descriptive Statistics
date: 2026-03-04
date-format: long
published-title: "Last updateded on"
abstract: "This chapter introduces basic descriptive statistics (e.g., mean, standard deviation etc.) and how to compute them using R. It is also available as a [Revealjs presentation](https://quarto.org/docs/presentations/revealjs/)."
footer: "[Back to riegersven.github.io](https://riegersven.github.io/library/working-with-data/descriptive-statistics.html)"
---

::: {.content-hidden when-profile="slides"}

## Revealjs Presentation {.unnumbered}

If you want to see the presentation in full screen click [here](../_slides/working-with-data/descriptive-statistics.html).

<div class="container">
  <iframe class="responsive-iframe" src="../_slides/working-with-data/descriptive-statistics.html"></iframe>
</div> 

:::

::: {.content-hidden unless-profile="slides"}

# Agenda

- [Preface: Data matrix](#data-matrix)
- [Descriptive statistics](#descr-descriptives)
- [Exercise](#descr-exercise)

::: footer
This is a [`revealjs`](https://revealjs.com/) presentation made with [Quarto](https://quarto.org/).
:::

:::



## Preface: Data matrix {#data-matrix}

Variables (e.g., characterisics), units (e.g., persons) and data (e.g., measurements) are often presented in matrix form. A matrix is a system of $n \cdot p$ quantities and looks like in the following:

::: {.fragment}

$$
\begin{bmatrix}
X_{11} & X_{12} & \cdots & X_{1p} \\
X_{21} & X_{22} & \cdots & X_{2p} \\
\vdots & \vdots &  & \vdots \\
X_{n1} & X_{n2} & \cdots & X_{np}
\end{bmatrix}
$$

:::

::: {.fragment}

- $n$ rows; 1 row is also known as a vector or row matrix
- $p$ columns; 1 column is also known as a vector or column matrix

:::

::: aside
see @eid2013
:::

::: {.content-hidden unless-profile="slides"}

# Descriptive statistics {#descr-descriptives}

:::



## Overview

:::: {.columns}
::: {.column width="49.5%"}

::: {.fragment}

- **Frequencies**
  - [Absolute](#abs-freq)
  - [Relative](#rel-freq)

:::

::: {.fragment}

- **Quantiles** (not covered)

:::

::: {.fragment}

- **Measures of central tendency**
  - [Mean](#descr-mean)
  - Weighted mean (not covered)
  - Weighted geometric mean (not covered)
  - [Median](#descr-median)
  - Mode (not covered)
  - ...

:::

::: 


::: {.column width="1%"}
:::

::: {.column width="49.5%"}

::: {.fragment}

- **Measures of variability**
  - [Standard deviation](#descr-sd)
  - [Variance](#descr-var)
  - [Range](#descr-range) (Minimum, Maximum)
  - Interquartile range (not covered)
  - Semi-interquartile range (not covered)
  - ...
:::

::: {.fragment}

- **Measures of shape**
  - Skewness (not covered)
  - Kurtosis (not covered)

:::

```{r}
#| label: norm-dis
#| eval: false
#| fig-height: 6
#| echo: false
ggplot(data.frame(x = c(-3, 3)), aes(x)) +
  stat_function(fun = dnorm,
                geom = "area",
                fill = "darkgrey",
                alpha = 1,
                args = list(mean = 0, sd = .5)) +
  stat_function(fun = dnorm,
                geom = "area",
                fill = "steelblue",
                alpha = .6,
                args = list(mean = 0, sd = 1)) +
  stat_function(fun = dnorm,
                geom = "area",
                fill = "#009E73",
                alpha = .6,
                args = list(mean = 1, sd = .5)) +
  stat_function(fun = sn::dsn,
                geom = "area",
                fill = "darkred",
                alpha = .6,
                args = list(xi = -2, omega = 1.25, alpha = 12)) +
  scale_x_continuous(breaks = seq(-3, 3, by = 1),
                     limits = c(-3, 3)) +
  theme_classic() +
  theme(axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank())
  

```

:::
::::

## Example data set

Consider the following 2 vectors within the example data set**

```{r}
#| label: exDat
#| code-line-numbers: true
#| echo: true

ex_dat <- data.frame(
  num_vec = c(1, 2, 5, 3, 8),
  chr_vec = c("low", "med", "low", "high", "high")
)
```

## Absolute Frequencies {#abs-freq}

::: {.callout-note appearance="simple" icon="false"}

Absolute frequencies refer to the numbers of a particular value or category appearing in a variable. It may be abbreviated with $n_j$ where $n$ is the number of a specific value/category $j$.

::: {.fragment}

+----------------+----------------------------+-------------------------------------+
| Category $j$   | Absolute Frequency ($n_j$) |                                     |
+:===============+:==========================:+:===================================:+
| low ($j=1$)    | 2                          |                                     |
+----------------+----------------------------+-------------------------------------+
| med ($j=2$)    | 1                          |                                     |
+----------------+----------------------------+-------------------------------------+
| high ($j=3$)   | 2                          |                                     |
+----------------+----------------------------+-------------------------------------+
| $\sum$         | $\sum_{j=1}^3n_j=n=5$      |                                     |
+----------------+----------------------------+-------------------------------------+

: Example Frequency table {tbl-colwidths="[20,40,40]" }

:::

:::

::: {.content-hidden unless-profile="slides"}

## Absolute Frequencies in R {#abs-freq-r}

:::

::: {.panel-tabset}

### base

```{r}
#| label: abs-freq-base-1
#| echo: true
#| output-location: column-fragment
with(ex_dat,
     table(chr_vec)) 
```


::: {.fragment}

An important argument (`useNA`) and another useful function (`addmargins()`)...

```{r}
#| label: abs-freq-base-2
#| echo: true
#| output-location: column-fragment
with(ex_dat,
     table(chr_vec, useNA = "always")) |>
  addmargins() 
```

:::

### dplyr

```{r}
#| label: abs-freq-dplyr
#| echo: true
#| output-location: column
ex_dat |>
  dplyr::group_by(chr_vec) |>
  dplyr::summarise(absFreq = dplyr::n())
```

:::

## Relative Frequencies {#rel-freq}

::: {.callout-note appearance="simple" icon="false"}

Relative frequencies refer to the proportion of a specific value or category relative to the total number of observations ($n$). 

$$
h_j=\frac{n_j}{n}
$$

::: {.fragment}

+----------------+----------------------------+-------------------------------------+
| Category $j$   | Absolute Frequency ($n_j$) | Relative Frequency ($h_j$)          |
+:===============+:==========================:+:===================================:+
| low ($j=1$)    | 2                          |  0.40                               |
+----------------+----------------------------+-------------------------------------+
| med ($j=2$)    | 1                          |  0.20                               |
+----------------+----------------------------+-------------------------------------+
| high ($j=3$)   | 2                          |  0.40                               |
+----------------+----------------------------+-------------------------------------+
| $\sum$         | $\sum_{j=1}^3n_j=n=5$      | $\sum_{j=1}^3h_j=1$                 |
+----------------+----------------------------+-------------------------------------+

: Example Frequency table {tbl-colwidths="[20,40,40]" }

:::


:::

::: {.content-hidden unless-profile="slides"}

## Relative Frequencies in R {#rel-freq-r}

:::

::: {.panel-tabset}

### base

```{r}
#| label: rel-freq-base-1
#| echo: true
#| output-location: column-fragment
with(ex_dat,
     table(chr_vec)/sum(table(ex_dat$chr_vec))) 
```


::: {.fragment}

Another useful function (`sprintf()`) to force 2 decimal and add %...

```{r}
#| label: rel-freq-base-2
#| echo: true
#| output-location: column-fragment
with(ex_dat,
     table(chr_vec)/sum(table(ex_dat$chr_vec))) |>
     (function(x) sprintf("%.2f%%", x*100))()

```

:::

### dplyr

```{r}
#| label: rel-freq-dplyr
#| echo: true
#| output-location: column
ex_dat |>
  dplyr::select(chr_vec) |>
  dplyr::group_by(chr_vec) |>
  dplyr::summarise(absFreq= dplyr::n()) |>
  dplyr::mutate(relFreq = absFreq/sum(absFreq)) 
```

:::

## Mean {#descr-mean}

::: {.callout-note appearance="simple" icon="false"}
The mean (or arithmetic mean, average) is the sum of a collection of numbers divided by the count of numbers in the collection. The formula is given in @eq-mean.

$$
\bar{x} = \frac{1}{n} \sum_{i=1}^n x_i=\frac{x_1+x_2+\dots+x_n}{n}
$${#eq-mean}

For example, consider a vector of numbers: $x = 1, 2, 5, 3, 8$

$$
\bar{x} = \frac{(1+2+5+3+8)}{5}=3.8
$$


If the underlying data is a sample (i.e., a subset of a population), it is called the sample mean.

:::

::: {.fragment}

How to calculate the mean in `R`?

```{r}
#| label: mean-r1
#| echo: true
#| output-location: column-fragment
#| code-line-numbers: false
with(ex_dat,
     mean(num_vec))
  
```

:::

## A brief note on missing data 

In `R` missing values/data are represented by the symbol `NA`. Most of the basic functions cannot deal *appropriately* with missing data.

<br>

:::: {.columns}

::: {.column width="49.5%"}

::: {.fragment}

To demonstrate this we create another example vector (`exVec2`). 

```{r}
#| label: mean-r2
#| echo: true
#| output-location: fragment
#| code-line-numbers: false
num_vec2 <- c(1, 2, 5, 3, 8, NA)
mean(num_vec2)
```

:::

:::

::: {.column width="1%"}

:::

::: {.column width="49.5%"}

::: {.fragment}

If there is missing data, we are required to set the argument `na.rm` to `TRUE`.

```{r}
#| label: mean-r3
#| echo: true
#| output-location: fragment
#| code-line-numbers: false
mean(num_vec2, na.rm = TRUE)
```

:::

:::

::::


<br>

::: {.callout-important appearance="simple" icon="true" .fragment}
Omitting or deleting missing values should–in most scenarios–be avoided altogether [@schafer2002; @enders2025]

:::

::: {.content-hidden unless-format="revealjs"}

::: footer
[More on missing data in R](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Missing-values)
:::

:::

## Median {#descr-median}

::: {.callout-note appearance="simple" icon="false"}
The median is the value separating the higher half from the lower half of a data sample, a population, or a probability distribution. For a data set, it may be thought of as "the middle" value. The formulas are given in @eq-med.

$$
Mdn = \widetilde{x} = 
\begin{cases}
x_{(n+1)/2}                 & \:\: \text{if } n  \text{ is odd} \\
(x_{n/2} + x_{(n/2)+1}) / 2 & \:\: \text{if } n  \text{ is even}
\end{cases}
$${#eq-med}

Consider again the vector of numbers: $x = 1, 2, 5, 3, 8$ with length $n = 5$. To calculate the median you need to first, order the the vector: $x = 1, 2, 3, 5, 8$ and then apply the corresponding formula (odd vs. even; here odd):

$$
\widetilde{x}=x_{\frac{(5+1)}{2}}=x_3 = 3
$$
:::


::: {.fragment}

How to calculate the median in `R`?

```{r}
#| label: med-r
#| echo: true
#| output-location: column-fragment
#| code-line-numbers: false
with(ex_dat,
     median(num_vec))
```

:::

## Variance {#descr-var}

::: {.callout-note appearance="simple" icon="false"}
The variance is the expectation of the squared deviation of a random variable from its mean. Usually it is distinguished between the population and the sample variance. The formula of the population variance is given in @eq-pop-var.

$$
VAR(X) = \sigma^2 = \frac{1}{N} \sum\limits_{i=1}^N (x_i - \mu)^2
$${#eq-pop-var}

The formula of the sample variance is given in @eq-samp-var.

$$
VAR(X) = s^2 = \frac{1}{n-1} \sum\limits_{i=1}^n (x_i - \bar{x})^2
$${#eq-samp-var}

Using again the vector $x = 1, 2, 5, 3, 8$, the sample variance is calculated as follows: 

$$
Var(X) =\frac{1}{4}((1-3.8)^2 + (2-3.8)^2 + (5-3.8)^2 + (3-3.8)^2 + (8-3.8)^2) = 7.7
$$

:::


::: {.fragment}

How to calculate the variance in `R`?

```{r}
#| label: var-r
#| echo: true
#| output-location: column-fragment 
#| code-line-numbers: false
with(ex_dat,
     var(num_vec))
```


:::



## Standard Deviation {#descr-sd}

::: {.callout-note appearance="simple" icon="false"}
The standard deviation is defined as the square root of the variance. Again, it is distinguished between the population and the sample variance. The formula of the population standard deviation is given in @eq-pop-sd.

$$
SD(X) = \sigma = \sqrt{\sigma^2}
$${#eq-pop-sd}

The formula of the population standard deviation is given in @eq-samp-sd.

$$
SD(X) = s = \sqrt{s^2}
$${#eq-samp-sd}

Recall the variance calculation from the previous [slide](#descr-var), the (sample) variance of the vector is $7.7$.

$$
SD(X) = \sqrt{7.7}=2.774887
$$

:::


::: {.fragment}

How to calculate the standard deviation in `R`?

```{r}
#| label: sd-r
#| echo: true
#| output-location: column-fragment
#| code-line-numbers: false
with(ex_dat,
     sd(num_vec))
```

:::

## Range{#descr-range}

::: {.callout-note appearance="simple" icon="false"}
The range of a vector is the difference between the largest (maximum) and the smallest (minimum) values/observations. 

$$
Range(x) = R =  x_{max}-x_{min}
$${#eq-range}

:::


::: {.fragment}

How to calculate the range in `R`?

```{r}
#| label: range-r-1
#| echo: true
#| output-location: column-fragment
#| code-line-numbers: false
with(ex_dat,
     range(num_vec))
```
:::

::: {.fragment}

Alternatively, calculate minimum and maximum separately...

```{r}
#| label: range-r-2
#| echo: true
#| output-location: column
#| code-line-numbers: false
with(ex_dat,{
     c(min(num_vec),
       max(num_vec))})
```



To compute the range apply @eq-range.

```{r}
#| label: range-r-3
#| echo: true
#| output-location: column
#| code-line-numbers: false
with(ex_dat,
     max(num_vec)-min(num_vec))
```

:::


::: {.content-hidden unless-profile="slides"}

# Exercise {#descr-exercise}

:::

## Put everything together 

::: {.callout-caution appearance="simple" icon="true" title="Exercise" collapse="show"}

Let us calculate descriptive statistics (e.g., mean, standard deviation, minimum and maximum) for multiple variables. For this exercise, we use a subset of the HSB dataset which is provided in the `merTools` package [@R-merTools] (for some details see [here](../example-data/data-hsb.qmd)):

```{r}
#| label: descr-imp-hsb
#| echo: true
#| eval: true
#| code-line-numbers: false
#| output-location: column-fragment
#install.packages("merTools")
dat <- merTools::hsb
head(dat, 10)
```

:::

::: {.content-hidden unless-profile="slides"}

## Solution {#descr-solution}

:::

::: {.panel-tabset}

### Step 1: Calculate descriptives 

A flexible approach would be to use the `apply()` function...

<!-- old # code-line-numbers: 1-4|6|7|8|9|10-15|16-17 -->

```{r}
#| label: state-var
#| echo: true
#| output-location: column


myVar <- c("Math achievement" = "mathach", # <1>
           "Gender" = "female", # <1> 
           "Socioeconomic status" = "ses", # <1>
           "Class size" = "size") # <1>

ex_descr <- apply( # <2>
  X = dat[,myVar], # <3>
  MARGIN = 2, # <4>
  FUN = function(x) { # <5>
    ret <- c( # <6>
             mean(x, na.rm = T), # <6>
             sd(x, na.rm = T), # <6>
             min(x, na.rm = T), # <6>
             max(x, na.rm = T) # <6>
             ) # <6>
    return(ret) # <7>
    }) # <7>
```

1. Create a (named) character vector of the variables by using the `c()` function.
2. Use the `apply` function to apply a or multiple function(s) on data (here: 4 columns).
3. The input is the dataset with the selected columns of interest (see 1.).
4. `MARGIN = 2` indicates that the function should be applied over columns.
5. Create the function that should be applied. Here we calculate the `mean()`, `sd()`, `min()` and `max()`.
6. Create a temporary `R` object, which should be later returned (here: the vector `ret`)
7. Return the temporary object and close functions.

::: aside
::: {.fragment}
There are also functions such as `colMeans()`, `colSums()`, `rowMeans()` and `rowSums()`.
:::
:::

### Step 2: Print the results

Print the results...

```{r}
#| label: print-descr-1
#| echo: true
#| code-line-numbers: false
#| output-location: column-fragment

print(ex_descr)
```


This is a weird format; variables should be in rows not columns. Transpose...


```{r}
#| label: print-descr-2
#| echo: true
#| output-location: column
ex_descr |>
  t() |>
  print()
```


Better, but still not really convincing...


### Step 3: Create results table

<!-- old code-line-numbers: 1|2-3|4|5|6|7-12|13-14|15-19|20-21 -->

```{r}
#| label: tbl-descr
#| tbl-cap: "Descriptive statistics"
#| tbl-colwidths: [40, 15, 15, 15, 15]
#| echo: true
#| eval: true
#| output-location: column
library(flextable)
ex_descr_table <- ex_descr |> #<1>
    t() |> #<2>
    as.data.frame() |> #<2>
    (\(d) cbind(names(myVar), d))() |>  #<3>
    flextable() |> #<4>
    theme_apa() |> #<5>
    set_header_labels( #<6>
      "names(myVar)" = "Variables", #<6>
      V1 = "Mean",  #<6>
      V2 = "SD", #<6>
      V3 = "Min", #<6>
      V4 = "Max") |> #<6>
    align(part = "body", align = "center") |> #<7>
    align(j = 1, part = "all", align = "left") |> #<7>
    add_footer_lines( #<8>
      as_paragraph(as_i("Note. "), #<8>
                   "This is a footnote.") #<8>
      ) |> #<8>
    align(align = "left", part = "footer") |> #<8>
    width(j = 1, width = 2, unit = "in") |> #<9>
    width(j = 2:5, width = 1, unit = "in")  #<9>

ex_descr_table #<10>
```

1. Take the results (here: `ex_descr` object)...
2. ...and `transpose` (i.e., using the `t()` function) and coerce it to a `data.frame` object (`as.data.frame()`) 
3. Use the so-called [lambda (or anonymous) function](<https://en.wikipedia.org/wiki/Anonymous_function>) to bind (using the `cbind()` function) the variable names as the first column to the dataset.
4. Apply the `flextable()` function.
5. Use the APA theme (`theme_apa()`).
6. Rename the column names (`set_header_labels()`).
7. Center body part of the table (`align()`).
8. Add a footnote (`add_footer_lines`) and align it to the left.
9. Change column width (`width`) to 2 resp. 1 inch.
10. Print the table.

::: {.content-visible when-profile="slides"}

::: notes
Lambda functions, also known as anonymous functions, are a feature in many programming languages, including R. In R, a lambda function is a function that is not bound to a name. This means that it can be created and used on the fly without defining it in the standard way with a name. Lambda functions are particularly useful for short, one-off functions that are used as arguments to other functions, like in apply family functions or in operations that involve the pipe operator (|>).
:::

:::



### Step 4: Export the table (optional)

If you are not using Quarto (what you should be doing, though), and you need to export the table, you can use the `save_as_docx()` function.

```{r}
#| label: docx-export
#| echo: true
#| eval: false
exDescrTab |>
  set_caption(caption = "Table X.\nDescriptive statistics") |>
  save_as_docx(path = "descr-tab.docx")
```

:::

## Descriptive statistics with the `psych` package {#descr-R}

- Alternatively, it is convenient to use additional R packages such as the `psych` package [@R-psych] to calculate descriptive statistics
 
- Here we use the `describe` function (with the `fast` argument set to `TRUE`) to calculate the descriptive statistics of all variables within the example data set

```{r}
#| label: tbl-descr-psych
#| tbl-cap: Descriptive statistics with the psych package
#| echo: true
#| eval: true
#| output-location: default

dat |>
  subset(select = -c(1)) |>
  psych::describe(fast = TRUE) |>
  flextable() |>
  colformat_double(digits = 2)
```


::: {.content-hidden when-profile="book"}

# Questions?

## References {#descr-ref-reveal}

:::


::: {.content-hidden when-profile="slides"}

## Some Questions

::: {.callout-note title="This is work in progress..."}

... and needs to be completed.

:::

```{r}
#| label: descr-q-1
#| echo: false
#| 
descr_q1 <- webexercises::mcq(
  opts = c("mean", "median", "skewness", "mode", answer = "skewness")
)
```

What is NOT a measure of central tendency? `r descr_q1`

```{r}
#| label: descr-q-2
#| echo: false

descr_q2 <- webexercises::mcq(
  opts = c("TRUE", answer = "FALSE")
)
```


Most functions in R can deal *appropriately* with missing data if the argument `na.rm` is set to `TRUE`. `r descr_q2`

:::