---
title: Graphical Visualization (with `ggplot2`)
date: 2026-03-04
date-format: long
published-title: "Last updateded on"
abstract: "This chapter provides an introduction to graphical visualization using the `ggplot2` package. It provides an overview of different types of plots and how to create them. It is also available as a [Revealjs presentation](https://quarto.org/docs/presentations/revealjs/)."
footer: "[Back to riegersven.github.io](https://riegersven.github.io/library/working-with-data/graphical-visualization.html)"
---


::: {.content-hidden when-profile="slides"}

## Revealjs Presentation {.unnumbered}

If you want to see the presentation in full screen click [here](../_slides/working-with-data/graphical-visualization.html).

<div class="container">
  <iframe class="responsive-iframe" src="../_slides/working-with-data/graphical-visualization.html"></iframe>
</div> 

:::

::: {.content-hidden unless-profile="slides"}

## Agenda for today

- [Purpose of Data Visualizing](#data-viz)

- [`ggplot2` package](#ggplot2-pkg) [@R-ggplot2]

- [The Different Diagrams](#overview-diagrams)

::: footer
This is a [`revealjs`](https://revealjs.com/) presentation made with [Quarto](https://quarto.org/).
:::

:::

## Purpose of Data Visualizing

::: {.content-hidden unless-profile="book"}

The purpose of data visualizing is versatile as it...

:::

::: {.incremental}

- facilitates understanding of data (i.e., makes data "accessible")

- reveals "hidden" structures, trends and patterns 

- identifies outliers

- helps to communicate results clearly (storytelling)

- ...

:::


## (Some) Example Diagrams

```{r}
#| label: load-ggplot2-0
#| echo: false
#| message: false
#| warning: false
library(ggplot2)
demoTxtSize <- 25
```


```{r}
#| label: make-plots
#| echo: false
#| warmorning: false

DemoBar <- ggplot(data = ggplot2::diamonds,
                   aes(x = cut)) +
                geom_bar(color = "black",
                         fill = "white") +
                labs(y = " ", x = " ") +
                ggtitle("Barplot") +
                theme_classic() +               
                 theme(text = element_text(size=demoTxtSize),
                       plot.title = element_text(hjust = 0.5))

DemoHist <- ggplot(data = ggplot2::diamonds,
                    aes(x = price)) +
                 geom_histogram(color = "black",
                                fill = "white") +
                 labs(y = " ", x = " ") +
                 ggtitle("Histogram") +
                 theme_classic() +               
                 theme(text = element_text(size=demoTxtSize),
                       plot.title = element_text(hjust = 0.5))

DemoScatter <- ggplot(data = ggplot2::diamonds,
                      aes(x = price, y = carat, color = cut)) +
                 geom_point() +
                 labs(y = "Carat", x = "Price") +
                 ggtitle("Scatterplot") +
                 theme_classic() +               
                 theme(text = element_text(size=demoTxtSize),
                       plot.title = element_text(hjust = 0.5))


set.seed(999)
time <- 100
lineDat <- data.frame(Date = seq(as.Date("2024-01-01"),
                      by = "day",
                      length.out = time), 
                Y = c(cumsum(rnorm(time)),
                      cumsum(rnorm(time)),
                      cumsum(rnorm(time))),
                Group = c(rep("Group 1", time),
                         rep("Group 2", time),
                         rep("Group 3", time)))
DemoLine <- ggplot(data = lineDat,
                      aes(x = Date, y = Y, color = Group)) +
                 geom_line() +
                 labs(y = "Y", x = "Date") +
                 ggtitle("Lineplot") +
                 theme_classic() +               
                 theme(text = element_text(size=demoTxtSize),
                       plot.title = element_text(hjust = 0.5))

DemoBoxplot <- ggplot(data = ggplot2::diamonds,
                      aes(x = cut, y = carat)) +
                 geom_boxplot() +
                 labs(y = "Carat", x = "Cut") +
                 ggtitle("Boxplot") +
                 theme_classic() +               
                 theme(text = element_text(size=demoTxtSize),
                       plot.title = element_text(hjust = 0.5))
```

::: {layout="[[1,1,1],[1,1,1]]"}

::: {.fragment fragment-index=1}

```{r}
#| label: demo-barplot
#| echo: false

DemoBar

```

:::

:::{.fragment fragment-index=2}

```{r}
#| label: demo-hist
#| echo: false
#| warning: false

DemoHist

```

:::

:::{.fragment fragment-index=3}
```{r}
#| label: demo-scatter
#| echo: false

DemoScatter
```

:::

:::{.fragment fragment-index=4}

```{r}
#| label: demo-line
#| echo: false

DemoLine
```

:::

::: {.fragment fragment-index=5}

```{r}
#| label: demo-boxplot
#| echo: false

DemoBoxplot
```

:::

:::

::: {.fragment fragment-index=6}
But there are more such as Dot plots, venn diagrams, maps, networks, & many more...!
:::






## The `ggplot2` package {#ggplot2-pkg}

::: {.content-hidden unless-profile="slides"}

### Preface

:::


First of all, there are other graphical R packages such as:

  - the `base` [@R-base] solution (e.g., `plot()` function)
  - `lattice` package [@lattice2008]
  - `plotly` package [@R-plotly]
  - ...


::: {.fragment}

The `ggplot2` package [@R-ggplot2] ...

:::

::: {.incremental}
- is based on the Grammar of Graphics [@wilkinson2005] [&rarr; graphs can be composed by independent components]{.fragment}

- has great flexibility (and add-on packages) [&rarr; nothing is *impossible*]{.fragment}

- has a broad community support (e.g., <https://stackoverflow.com/>), and AI tools help to generate `ggplot2` code 

- belongs to the `tidyverse` [@R-tidyverse] package collection
:::


::: aside
::: {.fragment}
see @wickham2011
:::
:::



### Key Components {#key-comp-ggplot}

::: {.fragment}

Every `ggplot` needs [3 components]{.fragment .highlight-blue} to produce a plot: 

:::

::: {.fragment}

```{r}
#| label: demo-ggplot-2
#| echo: true
#| eval: false
#| code-line-numbers: 1|2|3
#| output-location: column
ggplot(data = diamonds, #<1>
       aes(x = price, y = carat)) + #<2>
    geom_point() #<3>
    
```

1. the *data*,
2. the so-called *aesthetic mappings* (i.e., which variables from the dataset should be used and how should they be mapped in the plot), and
3. at least one layer that defines what kind of visualization is desired (e.g., `geom_point` for a scatter plot)

:::



::: {.callout-note appearance="default" icon="true" .fragment}

Whereas the *data* and the *aesthetic mappings* (`aes()`) are stated within the `ggplot()` function, the `geom_point` layer is added with a `+`.

:::




::: aside
::: {.fragment}
<https://ggplot2-book.org/getting-started>
:::
:::


### Getting Started with `ggplot2` 


::: {.fragment}

First, install and load the package.

```{r}
#| label: load-ggplot
#| echo: true
#| output-location: column
#install.packages("ggplot2")
library(ggplot2)
```

:::

::: {.fragment style="padding-top: 0.5em;"}

Setting a theme (optional). This can be done by using `theme_set()` function. There are some built-in themes such as `theme_minimal()`, `theme_bw()`, or `theme_classic()`.

```{r}
#| label: setTheme-classic
#| echo: true
#| code-line-numbers: false
theme_set(theme_classic())
```

:::

::: {.fragment style="padding-top: 0.5em;"}

Further, you can customize the theme by using `theme_update()` function. For example, you can change the text size, font family, and color.

```{r}
#| label: update-theme
#| echo: true
#| results: hide
# install.packages("extrafont") # this may take a while
extrafont::loadfonts() 
theme_update(text = element_text(
    size=25,
    family="Helvetica",
    color = "blue"))


```

:::

::: {.content-hidden unless-profile="book"}

::: {.callout-warning title="Or going crazy with a custom theme..." icon="true" collapse="true"}


```{r}
#| label: setTheme-fun
#| results: hide
#extrafont::loadfonts()
myTheme <- theme_set(theme_classic()) +
  theme(
    text = element_text(
        size = 25, 
        family = "Comic Sans MS",
        color = "pink"),
    panel.grid.major.y = element_line(
      color = "red",
      linewidth = 0.5,
      linetype = 2
    )
  )
```

:::

:::


::: {.content-hidden unless-profile="slides"}

### Example Data

:::

::: {.content-hidden unless-profile="book"}

## Example Data

For this exercise, we use (again) a subset of the HSB dataset which is provided in the `merTools` package [@R-merTools]:

::: {.column-margin}

For more details on the dataset, see Chapter [High School and Beyond](../example-data/data-hsb.qmd).

:::

:::

```{r}
#| label: import-data
#| echo: true
#| output-location: column-fragment
dat <- merTools::hsb
head(dat[,1:7])
```



## Mapping Plot Types to `geom_*` Layers

::: {.content-hidden unless-profile="book"}

To create different types of plots, you need to use different `geom_*` layers. Here are some examples:

:::

- bar chart &rarr; `geom_bar()` ([jump](#barplot))
- histogram &rarr; `geom_histogram()` ([jump](#hist))
- boxplot &rarr; `geom_boxplot()` ([jump](#boxplot))
- line chart &rarr; `geom_line()`
- scatter plot &rarr; `geom_point()` ([jump](#scatterplot))
- text labels &rarr; `geom_text()`
- ...



## Bar Chart {#barplot}

### `geom_bar` layer (basic)


```{r}
#| label: bar-0
#| echo: true
#| code-line-numbers: "|3"
#| code-annotations: true
#| output-location: column
#| results: hold
p_bp <- ggplot(data = dat, #<0>
                aes(x = factor(female))) + #<0>
            geom_bar() #<1>
```

0. Provide data and aesthetics mappings (see [Key Components](#key-comp-ggplot))
1. Add the `geom_bar()` layer



::: {.content-hidden unless-profile="slides"}

### `geom_bar` layer (basic, print)

:::

```{r}
#| label: bar-0p
#| echo: true
#| code-fold: true
#| code-line-numbers: false
print(p_bp)
```


### `geom_bar` layer (customized) 

```{r}
#| label: bar-1
#| echo: true
#| code-line-numbers: "3|4|5|6-8|9|"
#| output-location: column
p_bp_c <- ggplot(data = dat,
                aes(x = factor(female))) +
            geom_bar(color = "blue", #<1>
                     linewidth = 1.5, #<2>
                     fill = "pink") + #<3>
            scale_x_discrete( #<4>
               label = c("0" = "Male", #<4>
                         "1" = "Female")) + #<4>
            labs(y = "Count", x = "Gender") # <5>

```

1. `color` argument: Customize color of boarder
2. `linewidth` argument: Customize size of boarder
3. `fill` argument: Customize color of bars 
4. `label` argument  in `scale_x_discrete` function can be used to provide labels for categories
5. `labs` function:  Customize text on x- and y-axes

::: {.content-hidden unless-profile="slides"}

## `geom_bar` layer (customized, print) 

:::

```{r}
#| label: bar-1p
#| echo: true
#| code-fold: true
#| code-line-numbers: false
print(p_bp_c)
```


## Histogram {#hist}

### `geom_histogram` layer (basic)

```{r}
#| label: hist-0
#| code-line-numbers: "|3"
#| code-annotations: below
#| output-location: column

p_hist <- ggplot(data = dat, # <0>
                 aes(x = mathach)) + # <0>
            geom_histogram() # <1>
     
```

0. Provide data and aesthetics mappings (see key components [slide](#key-comp-ggplot))
1. Add the `geom_histogram()`

::: {.content-hidden unless-profile="slides"}

### `geom_histogram` layer (basic, print)

:::

```{r}
#| label: hist-0p
#| echo: true
#| code-line-numbers: false
#| code-fold: true
print(p_hist)
```



### `geom_histogram` layer (customized) {#hist-cust}

```{r}
#| label: hist-1
#| code-line-numbers: "|4|5|6|7|8-11|"
#| output-location: column
p_hist_c <- ggplot(data = dat,  
                   aes(x = mathach)) + 
                geom_histogram(  
                    color = "black", # <1>
                    fill = "white", # <2>
                    bins = 30, # <3>
                    binwidth = 1.5 # <4>
                    ) + 
                labs(y = "Count", # <5>
                     x = "Math Achievement", # <5>
                     title = "My nice histogram") # <5>       
```

1. `color`: argument for color of the boarders of the bars
2. `fill`: argument for the color of the bars (do not use it together with the fill argument in the `aes()`!) 
3. `bins` argument: Number of bins
4. `binwidth` argument: The width of the bins
5. `labs` layer: Provide names of axis and title


::: {.content-hidden unless-profile="slides"}

### `geom_histogram` layer (customized, print)

:::

```{r}
#| label: hist-1p
#| echo: true
#| code-fold: true
#| code-line-numbers: false
print(p_hist_c)
```


## Scatterplot {#scatterplot}

### `geom_point` layer (basic)

```{r}
#| label: sct-0
#| code-line-numbers: "|3"
#| output-location: column

p_scp <- ggplot(data = dat, # <0>
                aes(y = mathach, x = ses)) + # <0>
            geom_point() # <1>
     
```

0. Provide data and aesthetics mappings (see key components [slide](#key-comp-ggplot))
1. Add the `geom_point()`

::: {.content-hidden unless-profile="slides"}

### `geom_point` layer (basic, print)

:::

```{r}
#| label: sct-0p
#| echo: true
#| code-fold: true
#| code-line-numbers: false
print(p_scp)
```



### `geom_point` layer (customized)

```{r}
#| label: sct-1
#| code-line-numbers: "4|5|6|7|8-11|"
#| output-location: column
p_scp_c <- ggplot(data = dat,  
                  aes(y = mathach, x = ses)) + 
            geom_point(  
                color = "black", # <1>
                fill = "white", # <2>
                size = 3, # <3>
                shape = 21 # <4>
                ) + 
            labs(y = "Math Achievement", # <5>
                 x = "Socioeconomic status", # <5>
                 title = "A scatterplot") # <5>       
```

1. `color`: argument for color of the boarders of the points (if `fill` $\neq$ `NULL`).
2. `fill`: argument for the color of the bars (do not use it together with the fill argument in the `aes()`!). 
3. `size` argument: Size of the points.
4. `shape` argument: Shape of the points (for more see <https://ggplot2.tidyverse.org/articles/ggplot2-specs.html>).
5. `labs` layer: Provide names of axis and title.

::: {.content-hidden unless-profile="slides"}

### `geom_point` layer (customized, print)

:::

```{r}
#| label: sct-1p
#| echo: true
#| code-fold: true
#| code-line-numbers: false
print(p_scp_c)
```

### `geom_point` layer (colored by group)

```{r}
#| label: sct-2
#| code-line-numbers: "|3|4|5|6|7|8-9|10-13|"
#| output-location: column
p_scp_c2 <- ggplot(data = dat,  
                   aes(y = mathach, x = ses)) + 
                geom_point(aes(color = factor(female)), #<1>
                           alpha = .5, #<2>
                           size = 2.5) + #<3>
                scale_color_manual( #<4>
                  values = c("black", "red"), #<5>
                  labels = c("0" = "Male", #<6>
                             "1" = "Female")) + #<6>
                labs(y = "Math Achievement", # <7>
                     x = "Socioeconomic status", # <7>
                     color = "Gender", # <7>
                     title = "A customized scatterplot") # <7>       
```

1. Use another *aesthetic mappings* (`aes()` function) to provide information which points should be colored.
2. `alpha` argument: Refers to the opacity of the points.
3. `size` argument: Size of the points.
4. `scale_color_manual` layer: Customize the discrete `color` scale.
5. `values` argument: Change color
6. `labels` argument: Provide labels (alternatively, you could use the `factor()` function in advance)
7. `labs` layer: Provide names of axis, legend title, & title

::: {.content-hidden unless-profile="slides"}

### `geom_point` layer (colored by group, print)

:::

```{r}
#| label: sct-2p
#| echo: true
#| code-fold: true
#| code-line-numbers: false
print(p_scp_c2)
```

## Boxplot {#boxplot}

### `geom_boxplot` layer (basic)

```{r}
#| label: box-0
#| code-line-numbers: "|3"
#| output-location: column

p_boxp <- ggplot(data = dat, # <0>
                aes(y = mathach,  # <0>
                    x = factor(schtype))) + # <0>
            geom_boxplot() # <1>
     
```

0. Provide data and aesthetics mappings (see key components [slide](#key-comp-ggplot)).
1. Add the `geom_boxplot()`.

::: {.content-hidden unless-profile="slides"}

## `geom_boxplot` layer (basic, print)

:::

```{r}
#| label: box-0p
#| echo: true
#| code-fold: true
#| code-line-numbers: false
print(p_boxp)
```



### `geom_boxplot` layer (customized)

```{r}
#| label: box-1
#| code-line-numbers: "4|5|6|7|8-11|"
#| output-location: column
p_boxp_c <- ggplot(data = dat,  
                   aes(y = mathach,
                       x = factor(schtype))) + 
                geom_boxplot(  
                    color = "#00ff00", # <1>
                    fill = "salmon", # <2>
                    size = 3, # <3>
                    width = 0.5, # <4>
                    staplewidth = 0.25) + # <5>
                labs(y = "Math Achievement", # <6>
                     x = "School Type", # <6>
                     title = "A customized boxplot") # <6>       
```

1. `color`: argument for color of the boarders of the points (if `fill` $\neq$ `NULL`).
2. `fill`: argument for the color of the bars (do not use it together with the fill argument in the `aes()`!). 
3. `size` argument: Size boarders.
4. `width` argument: Controls width of the bars.
5. `staplewidth` argument: Controls the width of the staples.
6. `labs` layer: Provide names of axis and title.

::: {.content-hidden unless-profile="slides"}

### `geom_boxplot` layer (customized, print)

:::

```{r}
#| label: box-1p
#| echo: true
#| code-fold: true
#| code-line-numbers: false
print(p_boxp_c)
```


## More Adjustment Options

### Facetting plots

Facetting allows to split data into subsets and display them across different plots. This can be done with the `facet_wrap()` or `facet_grid()` functions. To demonstrate the `facet_grid()` option, I used the customized histogram (`p_hist_c`) which was generated [here](#hist-cust).

::: {.content-hidden when-profile="book"}

::: {.r-stack}

::: {.fragment}

```{r}
#| label: demo-facet-grid-row
#| code-line-numbers: "2"
#| output-location: column
p_hist_c +
     facet_grid(rows = vars(female)) +
     labs(title = paste(
          "Overwrite the title", 
          "to demonstrate facet_grid",
          "(row dimension)", 
          sep = "\n")
          )
```

:::

::: {.fragment}

```{r}
#| label: demo-facet-grid-col
#| code-line-numbers: "2"
#| output-location: column
p_hist_c +
     facet_grid(cols = vars(female)) +
     labs(title = paste(
          "Overwrite the title", 
          "to demonstrate facet_grid",
          "(column dimension)", 
          sep = "\n")
          )
```

:::

::: {.fragment}

```{r}
#| label: demo-facet-grid-col-row
#| code-line-numbers: "2-3"
#| output-location: column
p_hist_c +
     facet_grid(cols = vars(female),
                rows = vars(minority)) +
     labs(title = paste(
          "Overwrite the title", 
          "to demonstrate facet_grid",
          "(column and row dimension)", 
          sep = "\n")
          )
```

:::

:::

:::

::: {.content-hidden when-profile="slides"}

::: {.panel-tabset}

#### row dimension

```{r}
#| label: demo-facet-grid-row-book
#| code-line-numbers: "2"
#| output-location: column
p_hist_c +
     facet_grid(rows = vars(female)) +
     labs(title = paste(
          "Overwrite the title", 
          "to demonstrate facet_grid",
          "(row dimension)", 
          sep = "\n")
          )
```

#### column dimension

```{r}
#| label: demo-facet-grid-col-book
#| code-line-numbers: "2"
#| output-location: column
p_hist_c +
     facet_grid(cols = vars(female)) +
     labs(title = paste(
          "Overwrite the title", 
          "to demonstrate facet_grid",
          "(column dimension)", 
          sep = "\n")
          )
```


#### row and column dimension


```{r}
#| label: demo-facet-grid-col-row-book
#| code-line-numbers: "2-3"
#| output-location: column
p_hist_c +
     facet_grid(cols = vars(female),
                rows = vars(minority)) +
     labs(title = paste(
          "Overwrite the title", 
          "to demonstrate facet_grid",
          "(column and row dimension)", 
          sep = "\n")
          )
```

:::

:::

### Labeling factors

Labeling factors may helpful for data visualization (not necessarily for data analyses!), because `ggplot2` then directly access the labels. This can be done with the `levels` and `labels` arguments of the `factor()` function. 


::: {.fragment}

```{r}
#| label: label-factor-1
#| code-line-numbers: 1|2|3-4|
#| output-location: column
dat$Gender <- factor(dat$female, # <1>
                     levels = c(0, 1), # <2>
                     labels = c("male", # <3>
                                "female")) # <3>

```

1. Use `factor()` function and access variable.
2. Provide the original levels in the `levels` argument.
3. Provide "publication-ready" labels in the `labels` argument.

:::



::: {.fragment}

Check with the `str()` function.

```{r}
#| label: label-factor-2
#| code-line-numbers: false
#| output-location: column-fragment
str(dat$Gender) # <1>
```

:::

::: {.content-hidden when-profile="slides"}

::: {.panel-tabset}

### Original `female` variable

```{r}
#| label: demo-label-factor-1-book
#| code-fold: true
ggplot(data = dat,
       aes(y = mathach, x = factor(female))) +
     geom_boxplot(width = .2,
                  staplewidth = .2) +
     labs(title = "Boxplot with original female variable")

```

### Transformed `Gender` variable 

```{r}
#| label: demo-label-factor-2-book
#| code-fold: true
ggplot(data = dat,
       aes(y = mathach, x = Gender)) +
     geom_boxplot(width = .2,
                  staplewidth = .2) +
     labs(title = "Boxplot with transformed female variable Gender")
```

:::

:::

::: {.content-hidden when-profile="book"}

### Labeling factors: Original `female` variable

```{r}
#| label: demo-label-factor-1
#| code-fold: true
ggplot(data = dat,
       aes(y = mathach, x = factor(female))) +
     geom_boxplot(width = .2,
                  staplewidth = .2) +
     labs(title = "Boxplot with original female variable (that is declared as numeric)")

```

### Labeling factors: Transformed `Gender` variable 

```{r}
#| label: demo-label-factor-2
#| code-fold: true
ggplot(data = dat,
       aes(y = mathach, x = Gender)) +
     geom_boxplot(width = .2,
                  staplewidth = .2) +
     labs(title = "Boxplot with transformed female variable Gender (that is a factor)")
```

:::

::: {.content-hidden when-profile="book"}

# Questions?

## References {#graph-ref-reveal}

:::



::: {.content-hidden when-profile="slides"}

## Some Questions

::: {.callout-note title="This is work in progress..."}

... and needs to be completed.

:::

```{r}
#| label: graph-q-1
#| echo: false
#| 
graph_q1 <- webexercises::mcq(
  opts = c("1", "2", "4", answer = "3")
)
```

How many components are required to create a ggplot2 visualization? `r graph_q1`

```{r}
#| label: graph-q-2
#| echo: false

graph_q2 <- webexercises::mcq(
  opts = c("|>", "-", "%>%", answer = "+")
)
```

How do you add a new layer to a ggplot2 plot? `r graph_q2`

:::

